---
title: "Microbiome Analysis"
output: pdf_document
author: "Olajumoke Evangelina Owokotomo"
date: "4/30/2019"
header-includes:
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("Biobase")
library("grid")
library("ggplot2")
library("nlme")
library(stringr)
library(dplyr)
library(phyloseq)
library(tidyr)
```


## General information

* The dataset is a large phyloseq object saved as R datafile.
* The metadata has been converted to phyloseq format.
* OTU Table contains 2353 taxa and 992 samples for all time point (OTU table).
* The sample are Crohn's disease patients.
* Samples are randomized to take PlaceboIV or UstekinumabIV ni the trial.
* Taxonomic assignment of the OTU into 6 taxonomic ranks (Tax data) which are;
"Kingdom" "Phylum"  "Class"   "Order"   "Family"  "Genus".
* 83 information was obtained about the 992 samples which includes their age, sex, race , country, treatment information, CDAI etc. 
* The CDAI is the standard instrument for evaluating clinical symptoms and disease activity in Crohn's disease.
* Difference in a subject’s initial Crohn’s disease activity index (CDAI) at baseline and other time point $\Delta_4 - \Delta_0$, $\Delta_6 - \Delta_0$, $\Delta_8 - \Delta_0$, $\Delta_{22} - \Delta_0$.
* Placebo samples were coded as 0 while treated samples were coded as 1.



```{r first, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
load("C:/Users/lucp2490.LP2490-02/Desktop/EVANGELINA/PHD FOLDER/PHD/DEA/certifi_ori.rda")
OTU <- data.frame(certifi_ori@otu_table)
TAX <- data.frame(certifi_ori@tax_table)
SAMPEL <- data.frame(certifi_ori@sam_data)

SAMPEL$Treatment= ifelse(SAMPEL$TRTGR=="Placebo",0,1)
```